---
title: "Confidence in the Government as It Relates to Inflation"
author: "Katherine Rafeiro Jackson"
description: "This data analysis project provides insights on how the inflation rate in the United States affects the population's confidence in the executive branch of the federal government."
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include = FALSE}

options(width = 100)
library(tidyverse)
library(infer)
library(broom)

confidence <- read_csv("Data/GOV50-FinalProject-Data-GSS-csv-prep.csv")

inflation <- read_csv("Data/Inflation-FRED-Jan01.csv")

confidence <- confidence |>
  filter(!year %in% c(1972,1985))

inflation <- inflation |>
  filter(!year  %in% c(1960:1972)) |>
  rename(perc_inflation = inflation_cpi)

conf_infl <- confidence |>
  inner_join(inflation, by = "year")

conf_infl <- conf_infl |>
  mutate(confed_nb = case_when(
            confed == "HARDLY ANY" ~ 1,
            confed == "ONLY SOME" ~ 2,
            confed == "A GREAT DEAL" ~ 3
  )) 

```

# Introduction

#### Research Question
This data analysis project sought to answer the following question: How does the US population's confidence in the executive branch of the federal government change as the inflation rate changes?

#### Hypothesis
In starting this project, my hypothesis was that as the inflation rate rises, the confidence in the executive branch of the federal government falls. This hypothesis makes intuitive sense because of the fact that high rates of inflation are not desirable by the population. While the Federal Reserve System's (the Fed's), the body in charge of controlling inflation, decisions do not need to be approved by the executive branch of the federal government, the Fed still "must work within the framework of the overall objectives of economic and financial policy established by the government." The Fed is therefore often associated to the executive branch of the federal government, even though it is not directly under its control.

With that being said, I think that the effect of the inflation rate on the confidence in the executive branch of the federal government is small. This makes intuitive sense because I recognize that there are a multitude of other factors that influence confidence such as the very particular historical context and important world events, whose effect on confidence is difficult to quantify. 

#### Relevance
This research question is interesting and important because we are currently in 2022, a time where inflation in the United States is very high, when compared to historical levels. The dataset that I used from the FRED - more on the dataset below - does not include levels of inflation for 2022, but I still believe this research question is important because a government's response (or lack thereof) to inflation could have major consequences on many aspects of society, for example, voting patterns. 

# Data and Research Design
The research design of this project is cross-sectional.

#### Description of the Data Source
I used two data sources for this project, which I ultimately merged together using the year.

The first is the General Social Survey (GSS) which was the source of my dependent variable. The GSS is a survey that collects information and keeps a historical record of the concerns, experiences, attitudes, and practices of residents of the United States.

The second is the "Inflation, consumer prices for the United States" from the World Bank, which I retrieved from the Federal Reserve Economic Data (FRED), and which was the source of my independent variable.  The FRED is a public database with macroeconomic indicators maintained by the Research division of the Federal Reserve Bank of St. Louis.

#### Independent Variable
My independent variable is the inflation rate. In the data downloaded from the FRED, this represents the "inflation_cpi" variable and it is coded in the following way: 1 percent = 1.00 and 6 percent = 6.00. The data was originally released by World Bank. This international organization used the Laspeyres formula to calculate inflation. 

```{r inflation, echo=FALSE}

ggplot(data = conf_infl, mapping = aes(x = year, y = perc_inflation)) + 
  geom_line() + 
  labs(title = "US Inflation Over Time", x = "Year", y = "Inflation (%)", subtitle = "From 1973 to 2021")

```

#### Dependent Variable
My dependent variable is the confidence in the executive branch of the federal government. In the GSS, this represents the "confed" variable and it is mostly coded as HARDLY ANY, ONLY SOME, and A GREAT DEAL, which I've converted to numeric data for the purposes of this analysis. Indeed, HARDLY ANY is 1, ONLY SOME is 2, and A GREAT DEAL is 3.

``` {r work_in_progress, echo=FALSE, fig.width = 8}

confed_mean_table <- conf_infl |>
  group_by(year) |>
  summarize(confed_mean = mean(confed_nb, na.rm = TRUE))

confed_mean_table |>
  ggplot(mapping = aes(x = year, y = confed_mean)) +
  geom_line() + 
  labs(title = "Confidence in the Executive Branch of the Federal Government Over Time", x = "Year", y = "Confidence", subtitle = "From 1973 to 2021")

ggplot(data = confed_mean_table, mapping = aes(x = confed_mean)) + 
  geom_histogram(binwidth = 0.01, fill = "skyblue3", aes(y=..count../sum(..count..))) + 
  labs(title = "Confidence in the Executive Branch of the Federal Government Over Time", x = "Confidence in the Executive Branch of the Federal Government", y = "Proportion")

```

## Results
Before regressing the confidence outcome variable on the inflation rate, the bivariate relationship between these variables can be visualized with a scatter plot. The graph below plots the inflation on the x-axis, and the confidence in the executive branch of the federal government on the y-axis. The years that are associated with each point are also indicated.

```{r regression, echo=FALSE,  fig.width = 8}

##Regression of inflation and confidence in the executive branch of the federal government. 
confed_infl_mean <- conf_infl |> 
  group_by(year) |> 
  summarize(confed_mean = mean(confed_nb, na.rm = TRUE), 
            mean_infl = mean(perc_inflation, na.rm = TRUE))

graph <- ggplot(data = confed_infl_mean, mapping = aes(x = mean_infl, y = confed_mean, label = year)) +
  geom_point() +
  geom_text(check_overlap = TRUE, hjust = 0, nudge_x = 0.1) +
  geom_smooth(method = "lm", se = FALSE, color = "blue1") + 
  labs(x = "Inflation (%)", y = "Confidence", title = "Inflation and Confidence in the Executive Branch of the Federal Government",  subtitle = "From 1973 to 2021")
graph

```
The scatter plot above suggests that there exists a slight positive relationship between the inflation rate and individuals' confidence in the executive branch of the federal government. Next, regression analyses will offer more detailed information about this seemingly positive association.

#### Regression results
Results from two regression models are below. The first is a bivariate model, which regresses the confidence on the inflation. The second is a multivariate model, which regresses confidence on inflation and year. 

```{r regressions, echo=FALSE}

mod1 <- lm(confed_mean ~ mean_infl, data = confed_infl_mean)

mod2 <- lm(confed_mean ~ mean_infl + year, data = confed_infl_mean)

var_labels <- c(
  "mean_infl" = "Mean Inflation", 
  "year" = "Year"
)

modelsummary::modelsummary(list(mod1, mod2),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

The bivariate model suggests that on average, a 1 point (1%) increase in inflation is associated with a 0.006 point increase in confidence in the executive branch of the federal government. On the other hand, the multivariate model suggests that on average, a 1 point (1%) increase in inflation is associated with a -0.021 point decrease in confidence in the executive branch of the federal government. 

```{r get_p_value, echo=FALSE, results='hide'}

#Coefficient estimates
coefest_mod1 <- summary(mod1)$coefficients["mean_infl", "Estimate"]
coefest_mod2 <- summary(mod2)$coefficients["mean_infl", "Estimate"]
coefest_mod1 
coefest_mod2

#P-values
pvalue_mod1 <- summary(mod1)$coefficients["mean_infl", "Pr(>|t|)"]
pvalue_mod2 <- summary(mod2)$coefficients["mean_infl", "Pr(>|t|)"]
pvalue_mod1
pvalue_mod2

#Standard errors
coefse_mod1 <- summary(mod1)$coefficients["mean_infl", "Std. Error"]
coefse_mod2 <- summary(mod2)$coefficients["mean_infl", "Std. Error"]
coefse_mod1
coefse_mod2

#Probabilistic confidence intervals
ci95_mod1 <- c(
  coefest_mod1 -1.96*coefse_mod1,
  coefest_mod1 + 1.96*coefse_mod1)
ci95_mod1

ci95_mod2 <- c(
  coefest_mod2 - 1.96*coefse_mod2,
  coefest_mod2 + 1.96*coefse_mod2)
ci95_mod2

```
In the bivariate model, the association between inflation and the confidence in the executive branch of the federal government is not statistically significant at the test level α=0.05. Specifically, the p-value on the coefficient is 0.5246107. If hypothesizing that the association between inflation and confidence rates is non-zero, then the coefficient’s p-value of 0.52 would suggest there is a 52% probability that the estimated association observed here is as extreme and meaningless as any estimated association from the null distribution.

In the multivariate model, the association between inflation and the confidence in the executive branch of the federal government is statistically significant at the test level α=0.05. Specifically, the p-value on the coefficient is 0.03. If hypothesizing that the association between inflation and confidence rates is non-zero, then the coefficient’s p-value of 0.03 would suggest there is a 3% probability that the estimated association observed here is as extreme and meaningless as any estimated association from the null distribution.

If we assume that the sampling distribution of confidence rates is approximately normal, we can construct a 95% confidence interval. This confidence interval means that if we take many samples, the regression correlation that we estimate between the confidence rate and the inflation rate will be within the interval 95% of the time.

Confidence interval for bivariate model: -0.011 and 0.022. 

Confidence interval for multivariate model:  -0.038 and -0.003. 

The confidence intervals concur with what was discussed above; the bivariate model is not statistically significant (because the confidence interval include 0), but the multivariate model is (because the confidence interval excludes 0). 

#### Causality and Confounders
Our results could be causal given the fact that the inflation rate is a national macroeconomic statistic. Therefore, inflation can be treated as a "treatment" variable that acts on the population's confidence rates. 

Our multivariate model attempts to control for one possible confounder: the year. Other possible confounders that weren't included in the model but that could be of relevance are race and political affiliation, both of which are also measured by the GSS. 

## Conclusion
To conclude, the bivariate regression demonstrates that the association between inflation and the confidence in the executive branch of the federal government is slightly positive. However, it is not statistically significant. On the other hand, the multivariate regression - that takes into account the year - demonstrates that this association is negative and statistically significant given its p-value of 0.03 when compared to α=0.05.

The fact that GSS's measure for confidence was imprecise - it was on a scale of 1 (HARDLY ANY) to 3 (A GREAT DEAL) - was a limitation of this analysis. To improve, it would be great to find and use data that has more precise measurement's for an individual's confidence in the executive branch of the federal government. Knowing an individuals state of residence, and the corresponding inflation rate in each state would also allow for a more precise analysis. 

Another possible improvement could be to adjust the type of regression used. In fact, I think my data points are best mapped to a quadratic regression. This would hold, unless I can prove some of my data to be outliers, at which point they could be removed from the regression model. 

It would also be interesting to plot inflation with the confidence in other governmental institutions, i.e. Congress, Supreme Court, which are available in the same GSS dataset. In other words, I could keep the same explanatory variable, but vary the outcome variables with a similar analysis. 
